<!-- ---
title: 'Statistical Lab 1 (Matrix Algebra with R) '
author: "Jiwan Bhandari"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: no
vignette: |
  %\VignetteIndexEntry{Vignette Title} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---
-->

<div style="text-align:center;min-height:29.7cm;width:21cm;vertical-align:center;">
<h4 style="">STAT 755 Lab Report 2</h4>
<h4>Jiwan Bhandari</h4>
</div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Dataset
For this excerise we generate our own sample data.  We first generate $1000*2$ matrix *N*  of standard random normal variables. We then generate $1000*3$ matrix *C* using linear combination of the columns of *N*.

```{r echo =TRUE}
   set.seed(c(123,233))
   N <- data.frame(x=rnorm(1000,mean=0,sd=1),y=rnorm(1000,mean=0,sd=1))
   C <- data.frame(c1=N$x+N$y,c2=N$x-N$y,c3=2*N$x+3*N$y)
``` 

#### Generalized Variance 
The generalized variance  of a p-dimensional random vector variable X is defined as the determinant of its variance-covariance matrix.Below we calculate the generalized variance of C. 
```{r echo=TRUE}
    cov_mat <- cov(C)
    S <- det(cov_mat)
    print(round(S,3))
```
The generalized variance of matrix C is 0 which suggest that the observations in C aren't linearly independent. 
#### Spectral Decomposition To Find  Linearly Dependent Columns
Spectral Decomposition of a $k*k$symmetric matrix   $A$ allows us to represent $A$ in terms of it's eigen values and orthogonal eigen vectors.
$$A= \mathbf{\sum_{i=1}^{k}{\lambda_ie_ie_i'}} $$. Below do the spectral decompositon of variance-covariance matrix of $C$ and find the linearly dependent columns of $C$ based on the spectral decomposition. 
```{r pressure, echo=TRUE}
  E<- eigen(cov_mat)
  

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
